Modular Type Constructors

### 2014-09-07 att-esop15/att-esop15.tex
Added initial commit of ESOP15 and updated abstract

\documentclass[9pt,preprint]{sigplanconf}
\usepackage{amsmath,tabu}
\usepackage{cancel}
\usepackage{amssymb} 
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\newcommand{\copyleft}{\reflectbox{\sffamily\copyright}}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
%\usepackage{cite}
%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...
\input{macros-atlam}
\input{macros-catlam}
%\usepackage{sfmath}
\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

@@

\addtolength{\floatsep}{-8px}

\newcommand{\compresslist}{
  \vspace{-1em}
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}

@@

\usepackage{todonotes}

\lefthyphenmin=10
\sloppy

\newcommand{\F}[1]{{\sf #1}~}
\newcommand{\FF}[1]{{\sf #1}}
%\usepackage{mathptmx}
%\usepackage[T1]{fontenc}

\usepackage{microtype}
    \renewcommand{\topfraction}{0.9}  % max fraction of floats at top
    \renewcommand{\bottomfraction}{0.9} % max fraction of floats at bottom
\newcommand{\rulename}[1]{({\textsc{\textsl{#1}}})}
\def \TirNameStyle #1{\small\rulename{#1}}
\renewcommand{\MathparLineskip}{\lineskiplimit=.6\baselineskip\lineskip=.6\baselineskip plus .2\baselineskip}

\newcommand{\keyrule}{{\color{red}~$\star$}}
\newcommand{\minput}{}
\newcommand{\moutput}{^{\color{gray}{+}}}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{prop}{Proposition}

\begin{document}

\conferenceinfo{-}{-} 
\copyrightyear{-} 
\copyrightdata{[to be supplied]}

%\titlebanner{{\tt \textcolor{Red}{{\small Under Review -- distribute within CMU only.}}}}        % These are ignored unless
%\preprintfooter{Distribute within CMU only.}   % 'preprint' option specified.

\title{Modular Type Constructors}
\subtitle{Type System Fragments as Safely Composable Libraries}
%\subtitle{Conservatively Extending a Type System From Within}
\authorinfo{}{}{}
%\authorinfo{Cyrus Omar\and Nathan Fulton\and Jonathan Aldrich}
 %         {Carnegie Mellon University}
  %        {\{comar, nfulton, aldrich\}@cs.cmu.edu}

\maketitle
\begin{abstract} 
Abstraction providers sometimes need to directly introduce new types and operators into existing languages. 
Ideally, such typed language fragments could be separately defined and safely composed. 
Unfortunately, existing tools and techniques do not come equipped with modular reasoning principles comparable to those available for library-based embeddings packaged using a modern module system. Each combination of fragments is semantically its own dialect of the language and metatheoretic and compiler correctness results must be established monolithically. 
Recent work has started to address this problem, e.g. by showing how to modularly reason about concrete syntax. Our focus here is on safely composing separately defined type system fragments.  We organize each fragment around a type constructor, as is usual practice when describing type systems, and sidestep the difficulties of abstract syntax by statically delegating semantic control over a small, fixed abstract syntax in a type-directed manner to static logic associated with a relevant type constructor.
The paper is organized around a minimal calculus, @$\lambda$ that permits surprisingly practical examples. We establish several strong semantic guarantees, notably \emph{type safety}, \emph{stability of typing} under extension and \emph{conservativity}: 
that the \emph{type invariants} that a finite set of fragments maintain are conserved under  extension. 
This involves lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each fragment using a form of ``internal'' type abstraction. 
Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules. %We discuss how to optionally mechanize this reasoning in Coq.%We call embeddings constructed using this strengthened formulation of active type constructors \emph{active embeddings}. A strength of the technique is that it does not require mechanized specifications and proofs to make the guarantees above, but these can be constructed, using existing proof assistants, to increase confidence that an active type constructor definition adequately implements a specified fragment.%We introduce @$\lambda$, a simply-typed lambda calculus with simply-kinded type-level computation where new indexed type and operator constructors can be declared  from within. %It is structured in four layers: at the top-level, extension providers declare new indexed type constructors and term-level operator constructors. 
%Type-level functions associated with operator constructors define their static and dynamic semantics, the latter by translation to a fixed typed nternal language. By lifting compiler correctness techniques into the language, the ``actively typed semantics'' guarantees type safety. Going further, the semantics enforce an abstraction barrier at extension boundaries that ensures that extensions are mutually conservative (i.e. they do not weaken or interfere with one another, so that they can  be used together in any combination). 
%%Totality of the type-level language guarantees decidability of type checking and equality constraints on type indices makes type equivalence decidable. 
%We intend @$\lambda$ as a minimal foundation for future work on safe language-integrated extension mechanisms for typed programming languages, but it is already quite expressive. We demonstrate by showing how a conventional concrete syntax can be introduced by type-directed dispatch to Core @$\lambda$, then discuss a number of typed language fragments that can be \emph{actively embedded} by this mechanism as orthogonal ``libraries''. %(and discuss the sorts of fragments that, as yet, cannot).
\end{abstract}

%\category{D.3.2}{Programming Languages}{Language Classifications}[Extensible Languages]
%\category{D.3.4}{Programming Languages}{Processors}[Compilers]
%\category{F.3.1}{Logics \& Meanings of Programs}{Specifying and Verifying and Reasoning about Programs}[Specification Techniques]
%\keywords
%type-level computation, typed compilation
\section{Introduction}
Typed programming languages are often described in \emph{fragments}, each defining contributions to a language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, each introduced in a different chapter \cite{pfpl}. A language is then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that builds in partial function types,  polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed further below).
It is left implied that the metatheory developed separately is conserved when fragments like these are composed to form a language.

@@

\section{@$\lambda$, By Example}\label{overview}

\begin{figure}
\small
\hspace{-5px}\input{syntax-table-IL}
\caption{Syntax of {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$}, our internal language (IL). We write internal types and terms in blue when they may contain the indicated ``unquote'' forms, discussed later.}
\label{syntax-IL}
%\vspace{-10px}
\end{figure}

Programs in @$\lambda$ are written as \emph{external terms}, $e$. The concrete syntax of external terms is essentially conventional, as suggested by the left side of the example in Figure \ref{example}. It desugars (purely syntactically) to a subtantially more uniform abstract syntax, shown in Figure \ref{syntax-EL} and discussed as we go on. We review recent techniques that permit modularly introducing new desugarings like these in Sec. \ref{desugaring}.

The static and dynamic semantics are specified simultaneously by a \emph{bidirectionally typed translation semantics} targeting an \emph{internal language}, with terms $\iota$. The two key judgements have the form:  \[\esynX{e}{\st\moutput}{\iota\moutput} ~~~~~\text{and}~~~~~ \eanaX{e}{\st}{\iota\moutput}\]
\noindent

These are pronounced ``$e$ (synthesizes / analyzes against) type $\sigma$ with translation $\iota$ under typing context $\Upsilon$ and tycon context $\Phi$''. We indicate ``outputs'' when introducing judgement forms by a \emph{mode annotation}, shown; these are not part of the judgement's syntax. The rules, shown in Figure \ref{statics-EL}, will be discussed incrementally as we explain the example in Figure \ref{example}.  Bidirectional typechecking is also sometimes called \emph{local type inference} \cite{Pierce:2000:LTI:345099.345100}. %We will see how we leverage this bidirectionality when we discuss literal forms below. % The judgement is only well-defined for \emph{valid} tycon contexts, written judgementally as $\vdash \Phi$ and \emph{valid} typing contexts, $\vdash_\Phi \Upsilon$, which we also specify below.
 \begin{figure*}[t!]
\small
\hspace{-5px}\input{example-table}
%\\~\\
%$\begin{array}{rcl}
%e\cdot\conclbl{lbl}(\conclbl{lbl}_1=e_1, \cdots, \conclbl{lbl}_n=e_n)  :=  \etarg{(\conclbl{lbl}; \svar{list}n[\klbl]~\conclbl{lbl}_1~\cdots~\conclbl{lbl}_n)}{e}{e_1; \cdots; e_n}
%\end{array}$
\caption{An example external term, $e_{ex}$, in concrete syntax (left), desugared to abstract syntax (right), with static terms shown in green in examples (only). The notation ${\small \scolor{\desugar{...}}}$ stands for an embedding of the indicated ${\small \scolor{\conclbl{label}}}$, ${\small \scolor{\concrx{regular expression}}}$, ${\small \scolor{\concstr{string}}}$ or numeral into the SL, with derived kinds  $\klbl$, $\krx$,  $\kstr$ and $\knat$, not shown. The signatures of helper functions, e.g. ${\small \scolor{\svar{nil}}}$ and ${\small \scolor{\svar{list}n}}$, are shown in Figure \ref{helper-sigs}.}
\label{example}
\end{figure*}

\paragraph{Internal Language} @$\lambda$ relies on a \emph{typed internal language} with support for type abstraction. In this paper, we use {$\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\iunit}\,{\times}\,{+}\}$} (Figure \ref{syntax-IL}), though our results do not depend on this precise choice of internal type constructors, only on the general forms of the judgements. We assume the internal statics are specified in the standard way by judgements for internal type formation {$\iDelta~\iTheta \vdash \tau$}, typing context formation { $\iDelta \vdash \iGamma$} and type assignment {$\iDelta~\iGamma \vdash \iota : \tau\moutput$}. The internal dynamics are also a standard structural operational semantics with judgements {\small $\iota \mapsto \iota\moutput$} and {$\iota~\mathtt{val}$} (cf. \cite{pfpl}).

We also here define a syntax for simultaneous substitutions, of terms for variables, $\gamma$, and of types for polymorphic type variables, $\delta$, and assume standard judgements ensuring that these are valid with respect to a valid context, $\Delta \vdash \gamma : \Gamma$ and $\vdash \delta : \Delta$. We apply these substitutions to terms, types and typing contexts using the syntax $[\gamma]\iota$, $[\delta]\iota$, $[\delta]\tau$ and $[\delta]\Gamma$ (in some prior work, application of a substitution like this is written using a hatted form, e.g. $\hat\gamma(\iota)$; we intend the same semantics but use a notation more consistent with standard substitution, e.g. $[\iota/x]\iota'$). We will omit  leading $\emptyset$ and $\cdot$ in examples; in specifications, the former is used for  metatheoretic finite mappings, the latter for metatheoretic ordered lists.

This style of specification has some  similarities to the Harper-Stone \emph{typed elaboration semantics} for Standard ML \cite{Harper00atype-theoretic}. There, however, external and internal terms were governed by a common type system. In @$\lambda$, {internal types} classify only internal terms. 
This style may thus also be compared to specifications for the first stage of a type-directed compiler, e.g. the TIL compiler for Standard ML \cite{tarditi+:til-OLD}, here lifted ``one level up'' into the language. We will see how the techniques developed to reason about typed compilation are lifted into our semantics to compositionally reason about type safety, which requires that the translation of every well-typed external term is a well-typed internal term, below.

\paragraph{Kinds and Static Terms}
\begin{figure}[t!]
\small
\hspace{-5px}\input{syntax-table-EL}\\
\caption{Abstract syntax of the external language (EL). Terms in gray are technical devices with no corresponding concrete syntax.}
\label{syntax-EL}
\end{figure}
\begin{figure}[t!]
\small
\hspace{-4px}\input{syntax-table-SL}
\caption{Syntax of the static language (SL).}
\label{syntax-SL}
\end{figure}
The workhorse of our calculus is the \emph{static language}, which itself forms a typed lambda calculus where 
%External terms are classified by (external) \emph{types}. 
\emph{kinds}, $\kappa$, classify \emph{static terms}, $\sigma$.  The syntax of the SL is given in Figure \ref{syntax-SL}. We note that the core of the SL is a total functional programming language based on several standard fragments: total functions, quantification over kinds, inductive kinds (constrained by a positivity condition to prevent non-termination), and products and sums (these closely follow \cite{pfpl} so we again omit some details).

The kinding judgement, specified  in Figure \ref{statics-SL}, takes the form $\sofkX{\st}{\kappa\moutput}$, where $\kDelta$ and $\kGamma$ are analagous to $\iDelta$ and $\iGamma$ (and analagous well-formedness judgements are defined, also omitted). The natural number $n$ is used as a technical device in our metatheory to prevent terms of the form $\sana{n}{\st}$ and $\ssyn{n}$ from arising if not permitted; these forms, discussed later, never need to be written directly (they would have no corresponding concrete syntax), so $n$ can assumed 0 for all user code. $\kXi$ is described below. The normalization judgement (specified in Figure \ref{dynamics-SL}) takes the form $\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$, where $\memD$, $\memG$ and $\aCtx$ are also technical devices that will be described later; they too can be ignored from the perspective of user code. We write $\st \Downarrow \st'$ iff $\snorm{\st}{\emptyset}{\cdot}{{\rightharpoonup}; \emptyset; \emptyset}{\st'}{\emptyset}{\cdot}$. \emph{Static values} are normalized static terms. Normalization can also raise an error (to indicate a type error in an external term, or a problem in a tycon definition, as we will discuss), indicated by the judgement $\serrX{\st}$. We omit error propagation rules.%We will refer to the relevant rules as we proceed.

 %We will specify the statics (i.e. kinding) and dynamics (i.e. normalization) of the static language in Figures \ref{statics-SL} and \ref{dynamics-SL}.%The normalization judgement takes the form $\snorm{\st}{\memD}{\memG}{\aCtx}{\st\moutput}{\memD\moutput}{\memG\moutput}$.  %This core can be considered  a simply-typed dialect of a language like Coq (we explore this further in Sec. \ref{coq}).

\paragraph{External Types are Static Values}

External types (or simply \emph{types}) are {static values} of kind $\kty$.  %We will return to kinds $\kity$ and $\kitm$ below.
The form $\sty{c}{\st}$ introduces a type by ``applying'' a \emph{tycon}, $c$, to a static term, $\st$, called the \emph{type index}. There is one built-in tycon, $\rightharpoonup$, classifying possibly partial external functions, discussed below. All other types are constructed by tycons defined in the tycon context and identified by name. We write tycon names in small caps, e.g. $\tcvar{lprod}$, and use $\tc$ as a metavariable ranging over these (Figure \ref{syntax-TC}). We will return to the other introductory form for types, $\sotherty{m}{\tau}$, which also serves only as a technical device, later. %\todo{mention otherty here?}

Two user-defined types are constructed on line 1 of Figure \ref{example}: a labeled product type declaring a single field labeled $\conclbl{venue}$ of type $\tcvar{rstr}~\concrx{([A-Z]+) \digit\{4\}}$, which classifies \emph{regular strings}, known statically to be in the regular language of the indicated regular expression, which can contain nested parenthesized groups. %We write these indices using conventional concrete syntax. 
%The regular expression can contain parenthesized groups.% We will discuss the operations these provide below. %Although the concrete syntax resembles function application, in our calculus tycons are not themselves static terms.

\paragraph{Index Kinds} The rule \rulename{k-ty} in Figure \ref{statics-SL} requires that the type index must be of the \emph{index kind} of the tycon, given by the \emph{index kind context}, $\kXi$, a simple mapping from tycons to kinds (Figure \ref{syntax-TC}). The initial index kind context, $\kXi_0$, is always included, defining the index kind of ${\rightharpoonup}$ as $\kty \times \kty$. 
The example index kind context in Figure \ref{example-tcsigs}, $\kXi_\text{ex}$, specifies that the index kind of $\tcvar{lprod}$ is $\klist{\kprod{\klbl}{\kty}}$, and the  index kind of $\tcvar{rstr}$ is $\krx$. We assume $\klist{\kappa}$, $\klbl$ and $\krx$ are embedded into the static language in the standard way. The concrete syntax used on line 1 desugars to static terms of these kinds (this sugar could be defined modularly as \emph{kind-specific syntax} by adapting the technique described in Sec. \ref{desugaring} \cite{TSLs}). If a tycon context, $\Phi$, is valid, it  uniquely determines a valid index kind context, as specified  in Figure \ref{statics-TC} by the judgement $\tccok{\Phi}{\kXi\moutput}$, detailed below. In our example, $\tccok{\Phi_\text{ex}}{\kXi_\text{ex}}$.

\paragraph{Type-Level Computation}
The labeled product type on line 1 is bound to the static variable $\svar{V}$  in the remainder of the external term. This  desugars to the more general external construct $\eslet{\st}{\svar{x}}{e}$, on the right. The rules \rulename{syn-slet} and \rulename{ana-slet}, shown in Figure \ref{statics-EL}  (here, the former applies) begin by assigning a kind to $\sigma$ under the index kind context determined by $\Phi$, then substitute its value into the external term $e$ to continue with synthesis or analysis. Static variables, written in bold, are distinct from external and internal variables. The rule \rulename{n-ty} in Figure \ref{dynamics-SL} shows that normalizing $\sty{c}{\st}$ simply involves normalizing $\st$. Because types are static values but need not be written directly in normal form, this is  a form of \emph{type-level computation of higher kind}, though our SL has a more general role than a standard type-level language. %Type indices (e.g. common regular expressions) and, as we will shortly discuss, term indices can also be computed.%, so our use of a static language type-level computation.

\begin{figure}[t]
\small
\hspace{-5px}$\begin{array}{rcllcl}
\multicolumn{3}{l}{\mvlbl{tycons}~~~c~::=~{\rightharpoonup}\,~|~ \tc} & \multicolumn{3}{l}{\mvlbl{index kind contexts}}\\
\multicolumn{3}{l}{\mvlbl{tycon contexts}} & \kXi_0 & := & {\rightharpoonup}[\kty \times \kty]\\
\Phi & ::= & \cdot ~|~ \Phi, \phi & \kXi & ::= & \kXi_0 ~|~ \kXi, \tc[\kappa]\\
\multicolumn{3}{l}{\mvlbl{tycon defs}} & \multicolumn{3}{l}{\mvlbl{tycon sigs}}\\
\phi & ::= & \tcdef{\tc}{\psi}{\opdefs}~~~~& \psi & ::= & \tcsig{\kappa}{\opsigs}\\
\multicolumn{3}{l}{\mvlbl{opcon structures}} & \multicolumn{3}{l}{\mvlbl{opcon sigs}}\\
\opdefs & ::= & \keyw{rep} = \st  & \Omega & ::= & \cdot\\
 & ~|~ & \omega, \analitop{\st} & & ~|~ & \Omega, \litsig{\kappa}\\
 & ~|~ & \opdefs, \synop{\opname{op}}{\st} & & ~|~ & \opsigs, \opsig{\opname{op}}{\kappa}
\end{array}$
\caption{Syntax of tycons and opcons + sigs and contexts.}
\label{syntax-TC}
\end{figure}
%
\begin{figure}
\small
\hspace{-5px}$\begin{array}{l@{\hspace{2px}}c@{\hspace{2px}}l@{\hspace{10px}}l@{\hspace{2px}}c@{\hspace{2px}}l}
\Phi_\text{ex} & := & \phi_\text{rstr}, \phi_\text{lprod} & \kXi_\text{ex} & := &\kXi_0, \tcvar{rstr}[\krx], \tcvar{lprod}[\klist{\klbl\times\kty}]\\
\phi_\text{rstr} & & \text{\normalsize(Figure \ref{tc-rstr})} & \psi_\text{rstr} & := &\tcsig{\krx}{\Omega_\text{rstr}}\\
\phi_\text{lprod}& &\text{\normalsize(Figure \ref{tc-lprod})} & \psi_\text{lprod} &:=& \tcsig{\klist{\klbl \times \kty}}{\Omega_\text{lprod}}\\
\Omega_\text{rstr} & := & \multicolumn{4}{r}{\litsig{\kstr}, \opsig{\opname{concat}}{\kunit}, \opsig{\opname{\#}}{\keyw{Nat}}, \opsig{\opname{coerce}}{\krx}, \opsig{\opname{case}}{\kunit}, \opsig{\opname{!rc}}{\kunit}}\\
\Omega_\text{lprod} & := & \multicolumn{4}{r}{\litsig{\klist{\klbl}}, \opsig{\opname{\#}}{\klbl}, \opsig{\opname{with}}{\klist{\klbl}}, \opsig{\opname{!ri}}{\kunit}}
\end{array}$
\caption{The contexts and signatures for Figures \ref{tc-rstr} and \ref{tc-lprod}.}
\label{example-tcsigs}
\end{figure}

\paragraph{Type Constructor Contexts}

\begin{figure}[t]
\small
~\hfill{\fbox{$\tccok{\Phi}{\kXi\moutput}$}}\vspace{-20px}
\begin{mathpar}
\hspace{-45px}\inferrule[tcc-emp]{ }{\tccok{\cdot}{\kXi_0}}
~~~~~~~~~~~~
\inferrule[tcc-ext]{
  \tccok{\Phi}{\kXi}\\
  \tcok{\kXi}{\phi}{\tc}{\ktyidx}
}{\tccok{\Phi, \phi}{\kXi, \tc[\ktyidx]}}
\end{mathpar}
%
%
~\hfill{\fbox{$\tcok{\kXi}{\phi}{\tc\moutput}{\kappa\moutput}$}}\vspace{-20px}
%
%
\begin{mathpar}
\hspace{-65px}\inferrule[tcdef-ok]{
  \tc \notin \text{dom}(\kXi)~~~~
  \tcsigok{\psi}{\ktyidx}~~~~
  \ocok{\kXi}{\omega}{\tc}{\psi}
}{
  \tcok{\kXi}{\tcdef{\tc}{\psi}{\omega}}{\tc}{\ktyidx}
}
\end{mathpar}
~\hfill{\fbox{$\tcsigok{\psi}{\kappa\moutput}$}}\vspace{-20px}
\begin{mathpar}
\hspace{-50px}\inferrule[tcsig-ok]{
  \keq{\emptyset}{\ktyidx}\\
  \ocsigok{\Omega}
}{
  \tcsigok{\tcsig{\ktyidx}{\Omega}}{\ktyidx}
}
\end{mathpar}
~\hfill\fbox{$\keq{\kTheta\minput}{\kappa\minput}$}\vspace{-16px}\begin{mathpar}
\hspace{-50px}\inferrule[keq-k]{\kvar \in \kTheta}{\keq{\kTheta}{\kvar}}
~~~~~~
\inferrule[keq-ind]{\keq{\kTheta, \kvar}{\kappa}}{
  \keq{\kTheta}{\kmu{\kvar}{\kappa}}}
~~~~~~
\inferrule[keq-unit]{ }{\keq{\kTheta}{\kunit}}
\\
\inferrule[keq-prod]{
  \keq{\kTheta}{\kappa_1}~~~~
  \keq{\kTheta}{\kappa_2}
}{
  \keq{\kTheta}{\kprod{\kappa_1}{\kappa_2}}
}
~~~~~~
\inferrule[keq-sum]{
  \keq{\kTheta}{\kappa_1}~~~~
  \keq{\kTheta}{\kappa_2}
}{
  \keq{\kTheta}{\ksum{\kappa_1}{\kappa_2}}
}
~~~~~~
\inferrule[keq-ty]{ }{\keq{\kTheta}{\kty}}
\end{mathpar}
~\hfill\fbox{$\ocsigok{\Omega}$}\vspace{-18px}
\begin{mathpar}
\hspace{-20px}\inferrule[ocs-rep]{ }{\ocsigok{\cdot}}
~~~~
\inferrule[ocs-lit]{
  \keq{\emptyset}{\kappa}
}{
  \ocsigok{\litsig{\kappa}}
}
~~~~
\inferrule[ocs-targ]{
  \ocsigok{\Omega}~~~~
  \opname{op} \notin \text{dom}(\Omega)~~~~
  \keq{\emptyset}{\kappa}
}{
  \ocsigok{\Omega, \opsig{\opname{op}}{\kappa}}
}
\end{mathpar}
$\karg := (\kunit \rightarrow (\kty \times \kitm)) \times (\kty \rightarrow \kitm)$~\hfill\fbox{$\ocok{\kXi}{\omega}{\tc}{\psi}$}%\vspace{-18px}
\begin{mathpar}
\inferrule[oc-rep]{
  \sofkz{\emptyset}{\emptyset}{\kXi}{\stx{rep}}{\karrow{\ktyidx}{\kity}}
}{
  \ocok{\kXi}{\keyw{rep}=\stx{rep}}{\tc}{\tcsig{\ktyidx}{\cdot}}
}

\inferrule[oc-lit]{
  \ocok{\kXi}{\omega}{\tc}{\tcsig{\ktyidx}{\Omega}}\\
  \sofkz{\emptyset}{\emptyset}{\kXi, \tc[\ktyidx]}{\stx{def}}{{\ktyidx}\rightarrow{\kappa}\rightarrow{\kargs}\rightarrow {\kitm}}
}{
  \ocok{\kXi}{\omega, \analitop{\stx{def}}}{\tc}{\tcsig{\ktyidx}{\Omega, \litsig{\kappa}}}
}

\inferrule[oc-targ]{
  \ocok{\kXi}{\omega}{\tc}{\tcsig{\ktyidx}{\Omega}}\\
  \sofkz{\emptyset}{\emptyset}{\kXi, \tc[\ktyidx]}{\stx{def}}{{\ktyidx}\rightarrow\kitm\rightarrow{\kappa}\rightarrow{\kargs}\rightarrow{(\kty\times \kitm)}}\\
}{
  \ocok{\kXi}{\omega, \synop{\opname{op}}{\stx{def}}}{\tc}{\tcsig{\ktyidx}{\Omega, \opsig{\opname{op}}{\kappa}}}
}
\end{mathpar}
\caption{Type constructor kinding.}
\label{statics-TC}
\end{figure} A tycon context, $\Phi$, is a list of tycon definitions, $\phi$ (Figure \ref{syntax-TC}). Each tycon definition is annotated with a \emph{tycon signature}, $\psi$, which specifies the index kind of the tycon as well as an \emph{opcon signature}, $\Omega$, which relates to the \emph{opcon structure}, $\omega$, discussed below.  The tycons used in our example are defined in Figures \ref{tc-rstr} and \ref{tc-lprod}.  In Figure \ref{statics-TC}, the judgement $\tcok{\kXi}{\phi}{\tc\moutput}{\ktyidx\moutput}$ ensures that their names are unique, their signatures are valid, and that the opcon structure respects the opcon signature.%, ``returning'' the information needed to extend $\kXi$.

% \begin{figure}[t]
% \small
% $\begin{array}{rcl}
% \karg & := & (\kunit \rightarrow (\kty \times \kitm)) \times (\kty \rightarrow \kitm)\\
% %\svar{arity0} : \klist{\karg} \rightarrow \kunit & \svar{syn} : \karg \rightarrow \kty \times \kitm\\
% %\svar{arity1} : \klist{\karg} \rightarrow \karg & \svar{ana} : \karg \rightarrow \kty \rightarrow \kitm\\
% %\svar{arity2} : \klist{\karg} \rightarrow \karg \times \karg
% \end{array}$
% \caption{The external semantics exposes arguments as list of pairs of static functions for synthesis and analysis, respectively. We assume the two helper functions above in the examples.}
% \label{karg}
% \end{figure}
\begin{figure}[t]
\small
%\renewcommand{\slam}[3]{\lambda #2{:}#1.#3}
\hspace{-5px}$\begin{array}{l}
\tcdef{\tcvar{rstr}}{\psi_\text{rstr}}{\\
\quad\keyw{rep}=\scolorc\slam{\krx}{\svar{tyidx}}{\sqity{\qcolorc\imu{s}{\keyw{str}\times\keyw{list}[s]}\scolorc}\color{black}}\\
% ana lit
\quad \analitop{
  \scolorc\slam{\krx}{\svar{tyidx}}{\slam{\kstr}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{let}~\svar{aok} : \kunit = \svar{arity0}~\svar{args}~\keyw{in}~\cnl
\quad\quad \keyw{letpair}~(\svar{chars}, \svar{groups})  = \svar{rmatch}~\svar{tyidx}~\svar{tmidx}~\keyw{in}~\cnl
\quad\quad\quad \sqitm{\qcolorc\ifold{s}{\keyw{str}\times\keyw{list}[s]}{\cnl\qcolorc
\quad\quad\quad\quad (\quq{\scolorc\svar{str\_of\_Str}~\svar{chars}\qcolorc},\quq{\scolorc\svar{rstrs\_of\_Strs}~\svar{groups}\qcolorc})}\scolorc}}}}
}\color{black},\\
% syn concat
\quad \synop{\opname{concat}}{\scolorc\slam{\krx}{\svar{ttyidx}}{\lambda \svar{ttr}{:}\kitm.\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\cnl
\quad\quad \keyw{letpair}~(\svar{targ}, \svar{a})  =\svar{arity2}~\svar{args}~\keyw{in}~\keyw{letpair}~(\svar{aty}, \svar{atr})=\svar{syn}~\svar{a}~\keyw{in}\cnl
\quad\quad \stycase{\tcvar{rstr}}{\svar{aty}}{\svar{atyidx}}{(\sty{\tcvar{rstr}}{\svar{rconcat}~\svar{ttyidx}~\svar{atyidx}},\cnl
\quad\quad\quad \sqitm{\qcolorc rsconcat~\quq{\scolorc\svar{ttr}\qcolorc}~\quq{\scolorc\svar{atr}\qcolorc}\scolorc})}{\sraise{\kty \times \kitm}}}}}}\color{black},\\
% syn case
%\quad \synop{\opname{case}}{\slam{\krx}{\svar{ttyidx}}{\slam{\kitm}{\svar{ttr}}{\slam{\kunit}{\svar{tmidx}}{\slam{\kargs}{\svar{args}}{\\
%\quad\quad \svar{arity2}~\svar{args}~(\slam{\karg}{\svar{a1}}{\slam{\karg}{\svar{a2}}{\svar{syn}~\svar{a1}~\slam{\kty}{\svar{a1ty}}{\slam{\kitm}{\svar{a1tr}}{\\
%\quad\quad\quad \keyw{let}~\svar{hdty} : \kty =\sty{\tcvar{rstr}}{\svar{rxhd}~\svar{ttyidx}}~\keyw{in}\\
%\quad\quad\quad \keyw{let}~\svar{tlty} : \kty =\sty{\tcvar{rstr}}{\svar{rxtl}~\svar{ttyidx}}~\keyw{in}\\

>>>>

\documentclass{llncs}
\usepackage{amsmath}
\usepackage{llncsdoc}
\usepackage{amssymb} 
%\usepackage{amsthm}
\usepackage{ stmaryrd }
\usepackage{mathpartir}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\newcommand{\copyleft}{\reflectbox{\sffamily\copyright}}
%\input{macros-atlam}
\input{macros-catlam}
%\usepackage{cite}
%\renewcommand{\citepunct}{,\,} % IEEEtran wants to use ],\,[ for this but that looks dumb...

\renewcommand{\ttdefault}{txtt}
\usepackage{alltt}
\usepackage{listings}
\lstset{language=ML,
showstringspaces=false,
basicstyle=\ttfamily\footnotesize,
morekeywords={newcase,extends}}

@@

@@

\usepackage{todonotes}
\lefthyphenmin=7
\sloppy

\begin{document}
\mainmatter  % start of an individual contribution

% first the title is needed
\title{Modular Type Constructors}
\subtitle{Conservatively Composing Typed Language Fragments}
% a short form should be given in case it is too long for the running head
\titlerunning{Lecture Notes in Computer Science: Authors' Instructions}

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Cyrus Omar%
\and Jonathan Aldrich}
%
\authorrunning{Lecture Notes in Computer Science: Authors' Instructions}
% (feature abused for this document to repeat the title also on left hand pages)

% the affiliations are given next; don't give your e-mail address
% unless you accept that it will be published
\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
\texttt{\{comar,aldrich\}@cs.cmu.edu}
}

%
% NB: a more complex sample for affiliations and the mapping to the
% corresponding authors can be found in the file "llncs.dem"
% (search for the string "\mainmatter" where a contribution starts).
% "llncs.dem" accompanies the document class "llncs.cls".
%

\toctitle{Lecture Notes in Computer Science}
\tocauthor{Authors' Instructions}
\maketitle

\begin{abstract}
Abstraction providers sometimes need to extend a programming language with new base types and their corresponding operators. 
Ideally, such language fragments could be defined and reasoned about separately. Unfortunately, existing tools and techniques do not come equipped with strong modular reasoning principle -- fundamentally, %comparable to those available for library-based embeddings packaged using a modern module system. 
each combination of fragments becomes its own dialect of the language for which metatheoretic and compiler correctness results must be established monolithically. 
Recent work has started to address this problem, e.g. by showing how to modularly reason about desugarings atop a fixed type system. Our focus here is on safely composing type system fragments. We organize fragments around type constructors, as is usual practice when describing type systems, and sidestep the difficulties of composing abstract syntax (i.e. the expression problem)  by delegating control over a small, uniform abstract syntax in a type-directed manner to logic associated with a type constructor.
The paper is organized around a minimal calculus, @$\lambda$, that permits surprisingly practical examples. We establish several strong semantic guarantees, notably \emph{type safety}, \emph{stability of typing} under extension and \emph{conservativity}: 
that the \emph{type invariants} that a finite set of fragments maintain are conserved under  extension. 
This involves lifting typed compilation techniques into the semantics and enforcing an abstraction barrier around each fragment using a form of ``internal'' type abstraction. 
Conservativity then follows from classical parametricity results, so these \emph{modular type constructors} can be reasoned about like modules in an ML-like language. 
%\keywords{extensible languages; typed compilation; type-level computation; type abstraction}
\end{abstract}

\section{Introduction}
Typed programming languages are often described in \emph{fragments}, each defining contributions to a language's concrete syntax, abstract syntax, static semantics and dynamic semantics. 
In his textbook, Harper organizes fragments around type constructors, each introduced in a different chapter \cite{pfpl}. A language is then identified by a set of type constructors, e.g. $\mathcal{L}\{{\rightharpoonup}\,{\forall}\,{\mu}\,{\keyw{1}}\,{\times}\,{+}\}$ is the language that builds in partial function types,  polymorphic types, recursive types, nullary and binary product types and binary sum types (its syntax is shown in Figure \ref{syntax-IL}, discussed further below).
It is left implied that the metatheory developed separately is conserved when fragments like these are composed to form a language.

@@

% CANT GUARANTEE THAT SPECIFICATIONS ARE ACTUALLY DECIDABLE 
\bibliographystyle{abbrv}
\bibliography{../research}
\appendix
\section{Appendix}

\end{document}
